---
title: "Power analysis"
author: "James Galante"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Save Image for Debugging
```{r}
# Saving image for debugging
save.image("RDA_objects/visualize_power_results.rda")
message("Saved Image")
# stop("Manually Stopped Program after Saving Image")
```

### Goal
Exploration of power simulation results from the `r snakemake@wildcards$sample` dataset.

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
```

***

### Power across effect sizes
The statistical power to detect different effect sizes was computed by simulating UMI count data
and injecting the specified effect sizes (% decrease in expression). Here, the statistical power 
and the number of E-G pairs is at 80% power plotted for effect sizes.

```{r, message = FALSE}
# load results
results <- read_tsv(snakemake@input[[1]], progress = FALSE)
```

```{r, fig.height=4, fig.width=8}

# add power rank and percent of pairs
power <- results %>% 
  mutate(power_rank = seq_len(n()),
         power_pct = power_rank / n())

# plot power distribution
p1 <- ggplot(power, aes(x = power_pct, y = power, color = "darkblue")) +
  geom_hline(yintercept = 0.8, lty = "dashed") +
  geom_line(lwd = 1, color = "darkblue") +
  labs(title = "Power across tested pairs", y = "Power", x = "Tested perturbation-gene pairs",
       color = "Effect size") +
  scale_x_continuous(labels = scales::percent) +
  theme_bw() +
  theme(text = element_text(size = 13))

# filter for pairs with >= 0.8 power
power_80 <- filter(power, power >= 0.8)
# Calculate the percentage of pairs with at least 80% power
percent_80_power <- nrow(power_80) / nrow(power) * 100

# plot number of pairs at 80% power
p2 <- ggplot(power_80, aes(x = effect_size, fill = "darkblue")) +
  geom_bar(fill = "darkblue") +
  scale_y_continuous(limits = c(0, nrow(results))) +
  labs(x = "Effect size", y = "Tested perturbation-gene pairs",
       title = "80% Power",
       subtitle = sprintf("%s%% of Pairs", format(percent_80_power, digits = 2))) +
  theme_bw() +
  theme(legend.position = "none", text = element_text(size = 13))

# print plots as one figure
plot_grid(p2, p1, ncol = 2, rel_widths = c(0.33, 0.66))
```

***

### Power vs. gene expression and number of perturbed cells
Power is plotted as a function of gene expression levels and number of perturbed cells per
perturbation. For this, genes are binned into decentiles based on their average expression
(UMI/cell). For each bin, the average UMI/cell per gene is then calculated and used as the
expression level for genes in this bin. Similarly, the perturbations are grouped into quartiles
based on the number of cells per perturbation. The average statistical power per expression bin is
then plotted as a function of detected UMIs for each cells per perturbation quartile.

```{r}
# Define breakpoints and compute midpoints
breakpoints <- snakemake@params$num_cells_per_pert
midpoints <- c(0, (head(breakpoints, -1) + tail(breakpoints, -1)) / 2, Inf)

# Adjusting bins and adding a small jitter
set.seed(123)
power <- power %>%
  mutate(cells_quant = cut(mean_pert_cells,
                           breaks = midpoints,
                           labels = breakpoints,
                           include.lowest = TRUE, right = FALSE),
         avg_expr_jittered = average_expression_all_cells + runif(n = n(), min = -1e-6, max = 1e-6),
         expr_quant = cut(avg_expr_jittered, 
                          breaks = quantile(avg_expr_jittered, probs = seq(0, 1, 0.1), na.rm = TRUE),
                          include.lowest = TRUE))

# Aggregate data
avg_expr_quants <- power %>% 
  group_by(expr_quant) %>% 
  summarize(mean_expr = mean(average_expression_all_cells))

binned_power <- power %>%
  group_by(effect_size, cells_quant, expr_quant) %>% 
  summarize(mean_power = mean(power), .groups = "drop") %>%
  left_join(avg_expr_quants, by = "expr_quant")

# Plotting without line types and moving legend to the right
p <- ggplot(binned_power, aes(x = mean_expr, y = mean_power, color = cells_quant)) +
  geom_hline(yintercept = 0.8, lty = "dashed") +
  geom_line() +
  geom_point() +
  labs(title = "Power vs. gene expression", x = "Average UMI/cell per gene",
       y = "Power", color = "Combination") +
  scale_x_log10() +
  theme_bw() +
  theme(text = element_text(size = 13),
        legend.position = "right") +
  guides(color = guide_legend(title = "Number of\nperturbed cells"))

# If plotting in an R environment
print(p)
```


### Minimum number of UMIs for 80% power per number of pert cells

```{r, fig.height=4, fig.width=8}
# For the least number of UMIs for 80% power at each number of pert cells:
umi_80 <- power %>%
  filter(power >= 0.8) %>%
  group_by(cells_quant) %>%
  summarize(min_umi = min(average_expression_all_cells)) %>%
  ungroup()

# Plotting
p_umi_mean <- ggplot(umi_80, aes(x = cells_quant, y = min_umi, fill = cells_quant)) +
  geom_bar(stat = "identity") +
  labs(title = "Minimum UMIs for 80% Power", 
       x = "Number of Perturbed Cells (Not to Scale)", 
       y = "Minimum UMIs",
       subtitle = "Num Pert Cells = Mean of Sampling Distribution") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank()) 

# For the least number of UMIs for 80% power at each number of pert cells:
umi_80 <- power %>%
  filter(power >= 0.8) %>%
  group_by(mean_pert_cells) %>%
  summarize(min_umi = min(average_expression_all_cells)) %>%
  ungroup()

# Plotting
p_umi_sim <- ggplot(umi_80, aes(x = mean_pert_cells, y = min_umi)) +
  geom_line() +
  geom_point(color = "darkblue") +
  labs(title = "Minimum UMIs for 80% Power", 
       x = "Number of Perturbed Cells (To Scale)", 
       y = "Minimum UMIs",
       subtitle = "Num Pert Cells = Sampled Num from Poisson") +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

plot_grid(p_umi_mean, p_umi_sim, ncol = 2, rel_widths = c(0.5, 0.5))
```


### Percent of Pairs that Pass 80% at each Number of Perturbed Cells

```{r, fig.height=4, fig.width=8}
# Assuming 'power' is your data frame
pct_pairs_bin <- power %>%
  group_by(cells_quant) %>%
  summarise(percent_pass_80 = sum(power >= 0.8) / n()) %>%
  mutate(percent_pass_80 = percent_pass_80 * 100)

# Plotting
p_bin <- ggplot(pct_pairs_bin, aes(x = cells_quant, y = percent_pass_80, fill = cells_quant)) +
  geom_bar(stat = "identity") +
  labs(title = "Percent of Pairs with 80% Power", 
       x = "Number of Perturbed Cells (Not to Scale)", 
       y = "Percent of Pairs",
       subtitle = "Num Pert Cells = Mean of Sampling Distribution") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none") 
            

# Assuming 'power' is your data frame
pct_pairs_sim <- power %>%
  group_by(mean_pert_cells) %>%
  summarise(percent_pass_80 = sum(power >= 0.8) / n()) %>%
  mutate(percent_pass_80 = percent_pass_80 * 100)

# Plotting
p_sim <- ggplot(pct_pairs_sim, aes(x = mean_pert_cells, y = percent_pass_80)) +
  geom_line() +
  geom_point(color = "darkblue") +
  labs(title = "Percent of Pairs with 80% Power", 
       x = "Number of Perturbed Cells (To Scale)", 
       y = "Percent of Pairs",
       subtitle = "Num Pert Cells = Sampled Num from Poisson") +
  theme_bw()

plot_grid(p_bin, p_sim, ncol = 2, rel_widths = c(0.5, 0.5))
```

### Percent of Pairs that have 0 Power at each Number of Perturbed Cells

```{r, fig.height=4, fig.width=8}
# Assuming 'power' is your data frame
pct_pairs_bin <- power %>%
  group_by(cells_quant) %>%
  summarise(percent_0_power = sum(power == 0) / n()) %>%
  mutate(percent_0_power = percent_0_power * 100)

# Plotting
p_bin <- ggplot(pct_pairs_bin, aes(x = cells_quant, y = percent_0_power, fill = cells_quant)) +
  geom_bar(stat = "identity") +
  labs(title = "Percent of Pairs with 0% Power", 
       x = "Number of Perturbed Cells (Not to Scale)", 
       y = "Percent of Pairs",
       subtitle = "Num Pert Cells = Mean of Sampling Distribution") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = "none") 
 
pct_pairs_sim <- power %>%
  group_by(mean_pert_cells) %>%
  summarise(percent_0_power = sum(power == 0) / n()) %>%
  mutate(percent_0_power = percent_0_power * 100)
           
# Plotting
p_sim <- ggplot(pct_pairs_sim, aes(x = mean_pert_cells, y = percent_0_power)) +
  geom_line() +
  geom_point(color = "darkblue") +
  labs(title = "Percent of Pairs with 0% Power", 
       x = "Number of Perturbed Cells (To Scale)", 
       y = "Percent of Pairs",
       subtitle = "Num Pert Cells = Sampled Num from Poisson") +
  theme_bw()

plot_grid(p_bin, p_sim, ncol = 2, rel_widths = c(0.5, 0.5))
```


### Minimum number of cells to get 80% power by Mean UMIs

```{r}
# For the least number of perturbed cells for 80% power at each number of average expression:
umi_80 <- power %>%
  filter(power >= 0.8) %>%
  group_by(average_expression_all_cells) %>%
  summarize(min_pert_cells = min(mean_pert_cells)) %>%
  ungroup()

# Plotting
p_umi <- ggplot(umi_80, aes(x = average_expression_all_cells, y = min_pert_cells)) +
  geom_point(color = "darkblue") +
  geom_smooth(method = "nls", formula = y ~ a * exp(b * x), 
            method.args = list(start = list(a = max(umi_80$min_pert_cells), b = 0.1)), 
            color = "black", se = FALSE) +
  labs(title = "Minimum Perturbed Cells for 80% Power", 
       x = "Average Expression of Gene (UMIs)", 
       y = "Minimum Perturbed Cells",
       subtitle = "The Average Expression of a Gene is Calculated as the Mean UMIs from the Original Data") +
  theme_bw() +
  scale_x_log10() +
  theme(panel.grid.minor = element_blank())

print(p_umi)
```








